Predicting the Outcome of Patient-Provider Communication Sequences using Recurrent Neural Networks and Probabilistic Models

### 2018-01-11 paper-amia-summit-2018/amia_summit_2018.tex
camera ready version
update reviewer's comments and manuscript for camera-ready version

\textbf {Behavior code embeddings}. Representation of behavior codes was inspired by the recent success of word embeddings\cite{bengio2003neural}. Embedding is a representation of an object in low-dimensional space using a real-valued vector. In our study, embeddings of behavior codes were obtained as a by-product of training LSTM and GRU after feeding one-hot vectors as a representation of behavior codes as input to these RNNs. Behavior code embeddings have the property of representing similar codes with the vectors that are close to each other in low-dimensional space. Figure~\ref{fig:code_embedding} illustrates the MYSCOPE code embeddings visualized in 2-dimensional space by t-SNE \cite{maaten2008visualizing}. It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc. formed a cluster in the left part of Figure~\ref{fig:code_embedding}. The nearest neighbors of CQECML+ are highlighted by different color intensity (i.e. OQECML+ being more purple indicates that it is more similar to CQECML+). The right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc. It is interesting that the behaviors intended to elicit CHT+/CML+ group together, whereas the ones intended to elicit CHT-/CML- also group together and are located in the opposite regions of semantic space.

@@

\textbf {Recurrent Neural Networks (RNN)} are a class of neural networks that have an internal memory, which makes them particularly suitable for processing sequences of observations. The ability of RNNs to capture long-term dependencies and remember past observations for predicting future observations is their main advantage over MCs and HMMs. These features are very useful in analysis of motivational interviews, in which any behavior observed at a particular point in the interview may be indicative of other behaviors that are observed later. In order to mitigate the vanishing gradient problem of earlier versions of RNN \cite{bengio1993problem}, Hochreiter et al.\cite{hochreiter1997long} proposed Long Short Term Memory networks (LSTM). There are several variants of LSTM model, among which the most notable one is the Gated Recurrent Unit\cite{cho2014properties} (GRU). GRUs are simpler than LSTMs and have been shown to be effective for a variety of Natural Language Processing tasks \cite{cho2014properties}. GRU is formally defined as follows:
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\label{eq:resetgru}
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\label{eq:candidategru}
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
In Eq.~\ref{eq:firstgru}-\ref{eq:lastgru}, $\sigma$ corresponds to sigmoid function and $\odot$ designates an element-wise product. The update gate $z_t$ and reset gate $r_t$ at time step $t$ are computed by the Eq.~(\ref{eq:firstgru}) and~(\ref{eq:resetgru}), where $W_z$, $W_r$, $W_h$, $U_z$, $U_r$, $U_h$ are the weight matrices and $b_z$, $b_h$ and $b_r$ are bias vectors. The activation $h_t$ of the GRU at time $t$ is a linear combination of previous activation $h_{t-1}$ and the candidate activation $\tilde h_t$, which is represented by Eq.~(\ref{eq:lastgru}) and~(\ref{eq:candidategru}).

>>>>

\textbf {Behavior code embeddings}. Representation of behavior codes was inspired by the recent success of word embeddings\cite{bengio2003neural,mikolov2013distributed,pennington2014glove}. Embedding is a representation of an object in low-dimensional space using a real-valued vector. In our study, embeddings of behavior codes were obtained as a by-product of training LSTM and GRU after feeding one-hot vectors as a representation of behavior codes as input to these RNNs. Behavior code embeddings have the property of representing similar codes with the vectors that are close to each other in low-dimensional space. Figure~\ref{fig:code_embedding} illustrates the MYSCOPE code embeddings visualized in 2-dimensional space by t-SNE \cite{maaten2008visualizing}. It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc. formed a cluster in the left part of Figure~\ref{fig:code_embedding}. The nearest neighbors of CQECML+ are highlighted by different color intensity (i.e. OQECML+ being more purple indicates that it is more similar to CQECML+). The right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc. It is interesting that the behaviors intended to elicit CHT+/CML+ group together, whereas the ones intended to elicit CHT-/CML- also group together and are located in the opposite regions of semantic space.

@@

\textbf {Recurrent Neural Networks (RNN)} are a class of neural networks that have an internal memory, which makes them particularly suitable for processing sequences of observations. The ability of RNNs to capture long-term dependencies and remember past observations for predicting future observations is their main advantage over MCs and HMMs. These features are very useful in analysis of motivational interviews, in which any behavior observed at a particular point in the interview may be indicative of other behaviors that are observed later. In order to mitigate the vanishing gradient problem of earlier versions of RNN \cite{bengio1993problem}, Hochreiter et al.\cite{hochreiter1997long} proposed Long Short Term Memory networks (LSTM). There are several variants of LSTM model, among which the most notable one is the Gated Recurrent Unit\cite{cho2014properties} (GRU). GRUs are simpler than LSTMs and have been shown to be effective for a variety of Natural Language Processing tasks \cite{cho2014properties}. GRU is formally defined as follows:
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\label{eq:resetgru}
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\label{eq:candidategru}
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
In Eq.~\ref{eq:firstgru}-\ref{eq:lastgru}, $\sigma$ corresponds to sigmoid function and $\odot$ designates an element-wise product. The update gate $z_t$ and reset gate $r_t$ at time step $t$ are computed by the Eq.~(\ref{eq:firstgru}) and~(\ref{eq:resetgru}), where $W_z$, $W_r$, $W_h$, $U_z$, $U_r$, $U_h$ are the weight matrices and $b_z$, $b_h$ and $b_r$ are bias vectors. The activation $h_t$ of the GRU at time $t$ is a linear combination of the previous activation $h_{t-1}$ and the candidate activation $\tilde h_t$, which is represented by Eq.~(\ref{eq:lastgru}) and~(\ref{eq:candidategru}).
